---
title: "Estimating the Quality of BEAGLE Imputation with Masked Data"
author: "Gennady Khvorykh, [inzilico.com](http://inzilico.com)"
date: "`r Sys.Date()`"
output:
      rmarkdown::html_vignette:
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Estimating the Quality of BEAGLE Imputation with Masked Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Motivation
-----------

You have genotyped your samples on a SNP array and imputed the missing
genotypes with [BEAGLE](https://faculty.washington.edu/browning/beagle/beagle.html).
How good is the imputation? The internal metrics printed by BEAGLE (or stored
in the INFO field of the output) are estimates. The masked data analysis gives
you an empirical answer: hide a small proportion of the known genotypes,
impute them as if they were missing, and compare with the truth.

This vignette walks through the whole workflow with VCF files. BEAGLE itself
is not run inside the vignette; instead, the imputation is simulated by
copying the true genotypes back with a few errors introduced. Replace the
simulation with your real BEAGLE runs when you work with your data — the
shell commands are given in place.

The data
--------

We use a small VCF file with 20 individuals genotyped at 200 SNPs, which
ships with the package.

```{r}
library(imputeqc)
vcf_file <- system.file("extdata", "beagle.vcf", package = "imputeqc")
x <- ReadVCF(vcf_file)
```

`ReadVCF()` returns the VCF object, the haplotypes as strings, and the sample
and marker ids read from the file:

```{r}
length(x$haps)      # 2N haplotypes
nchar(x$haps[1])    # of M markers
head(x$samples, 3)
head(x$markers, 3)
```

Masking
-------

We hide 5% of genotypes at each locus and repeat this three times — with
only 20 samples, 1% would round to zero genotypes per locus; on real data
1% is the usual choice. The masks of one set are disjoint: together they
hide about 15% of the available genotypes, each hidden exactly once. Call
`set.seed()` beforehand to make the masks reproducible.

```{r}
set.seed(42)
masks <- GenerateMaskSet(x$haps, n = 3, p = 0.05,
                         samples = x$samples, markers = x$markers)
saveRDS(masks, "masks.RDS")
```

The ids passed via `samples` and `markers` are attached to the masks as
dimnames, make them self-describing, and enable the consistency checks of
`EstimateQuality()`.

```{r}
dim(masks[[1]])
sum(masks[[1]])          # hidden genotypes in the first mask
masks[[1]][1:4, 1:4]
```

The hidden genotypes can be exported for use outside R:

```{r}
WriteMaskSet(masks, "masks.tsv")
head(read.table("masks.tsv", header = TRUE))
```

`ApplyMasks()` writes one VCF per mask with the hidden genotypes set to
`./.`:

```{r}
ApplyMasks(x$haps, masks, pref = "masked", vcf = x$vcf)
```

Imputation
----------

In a real analysis you would run BEAGLE on every masked file, for example:

```{r, engine = "bash", eval = FALSE}
java -jar beagle.jar gt=masked.m1.vcf ref=reference.vcf.gz out=imputed.m1
java -jar beagle.jar gt=masked.m2.vcf ref=reference.vcf.gz out=imputed.m2
java -jar beagle.jar gt=masked.m3.vcf ref=reference.vcf.gz out=imputed.m3
```

**Important.** When imputing with a reference panel, the output contains all
reference markers, not only yours. Reduce every output back to the markers of
the original file before scoring, for example with bcftools:

```{r, engine = "bash", eval = FALSE}
bcftools view -T <(bcftools query -f '%CHROM\t%POS\t%REF\t%ALT\n' beagle.vcf) \
  -Oz -o imputed.m1.check.vcf.gz imputed.m1.vcf.gz
```

`EstimateQuality()` checks the dimensions, and, for VCF files, the sample and
marker ids, so a mis-aligned pair of files is reported instead of silently
producing nonsense.

Here we cannot run BEAGLE, so we stand in for it with a simulation: the
imputed file carries the true genotypes back, with a small fraction of the
hidden genotypes corrupted and one left missing. The essential contract is
simple — the imputed files must be VCFs (or fastPHASE files) with the same
samples and markers as the original file, carrying the imputed genotypes.

The files used below were produced by
`inst/extdata/make_beagle_vignette.R` of the package source, which performs
exactly this simulation with a 2% error rate on the same masks:

```{r}
imputed <- system.file("extdata", sprintf("beagle_imputed.m%d.vcf", 1:3),
                       package = "imputeqc")
imputed
```

Quality
-------

```{r}
eq <- EstimateQuality(origin = vcf_file, masks = "masks.RDS",
                      imputed = imputed, id = "beagle")
eq
```

The columns are the genotype discordance (the fraction of hidden genotypes
whose allele pair differs from the truth), the allele discordance, the number
of hidden genotypes, the number of them the imputer left missing, and the
ordinal number of the mask.

The discordance is usually strongly dependent on the minor allele frequency.
Pass `af_bins` to estimate it per stratum of the original data:

```{r}
eqb <- EstimateQuality(origin = vcf_file, masks = "masks.RDS",
                       imputed = imputed, id = "beagle",
                       af_bins = c(0, 0.01, 0.05, 0.1, 0.5))
eqb
```

Finally, `PlotDiscordance()` visualises the results of several experiments,
for example imputations made with different parameters:

```{r, fig.width = 6, fig.height = 4}
write.table(eqb, "discordance.tsv", sep = "\t", row.names = FALSE)
p <- PlotDiscordance("discordance.tsv", tl = "BEAGLE imputation")
print(p)
```

Session info
------------

```{r}
sessionInfo()
```
